---
title: "CCLHDME R Introduction"
output: 
  html_document:
    theme: united
date: "2022-07-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# clean with CSS


```

[clean/short/simple by Tuesday mid-day]

[Michael add info about SECURITY...]

[clean up our own stuff for easy installation -- email simple guide before meeting]

[We should revert our RStudios back to the default settings]


###  This R markdown document is a guide and tool for the workshop, and a way to share learning materials and other resources

###  The learning approach is watching leader live demonstrations, then participants doing the same things live.  Staring with immediately using R and R Studio as basically as possible for practical things, then cycle "backwards" for some concepts and options, then applying/using those, and repeat. And, pointing out and linking to additional learning resources and tools at each stage. Sound good? 



[Jaspo]

* what is R
* installing R
* Visit this website https://www.r-project.org/ and download the latest version of R suitable for your computer.
* & install R


* what is R Studio
* installing R Studio
* Visit this website https://rstudio.com/products/rstudio/download/ and download the latest free Desktop version of RStudio suitable for your computer.
* & install R Studio

[15 minutes]


* rstudio screen shot here

* walk through / overview R studio (and repeatedly return to this ...  Environment; Help, Chart, View, later Git)


[end Jaspo]



@ script 1 (console and script)


@ script 2 (console, then in script file)

  
### what are R packages

#### installing R packages



@ script 3 (script file)

* install a couple packages
* read some data
* wrangle some data
* ggplot some data



# Concepts and options for Session 1


* using R as a calculator
* script files
* simplest **objects** in R
  * vectors of length 1 (atomic vectors) 
  * vectors
  * data types
    * numeric
    * character
    * logical (Boolean)

* math with simple objects 

* R (mathematical) operators


* base R
  * indexing
  * charts
  
  
* reading data

* using a script file to save code and start a "program"

* looking at that data
  * names()
  * table()
  * plot with base R
  * etc.

* loading packages



* tidyverse
  * dplyr()
    * verbs - [find best reference resource for this]
    * pipes
  * ggplot2()
  * readr()
  * lubridate()
* using an R Project (?)
* thousands of other packages
  * summarytools()  - easy frequencies, "cross-tables", and summary of a data.frame
  * outbreaks() - line-listed data from a bunch of real outbreaks
  * epikit() or epitools() or something else 





joins (above or below)





## Some important concepts and options - Future sessions

* options for installing packages
* working with paths (?)

* data type precidence

*  Rules and best practices for naming objects

* more R object types
  *  data.frames and tibbles
  *  lists 
  *  class()
* functions
* markdown
  * https://skylab.cdph.ca.gov/communityBurden/_w_21945fed/xMDA/2020_Excess_Mortality.html
* ifelse() and other logical evaluation
* tables
* advanced data manipulation
* epidemiologic statistics and modeling
* shiny
* GIS
* shiny with GIS
* thousands of other packages
  * reading data
    * EXCELL
     * SAS  
  * Shiny - interactive visualizations and web sites
  * visual display
  * epidemiology
  * disease surveillance
  * outbreaks
  * GIS
    * tmap()
  * data analysis and modeling
  * tables
* GIT



Some important concepts and options - Future sessions
* apply and loops












```{r}
height <- rnorm(n=100, mean= 60, sd = 6)
weight <- height*1.2 +  rnorm(n=100, mean= 10, sd = 6)
hist(height, breaks=50)
plot(height, weight)
```



[Jaspreet and michael both show finding info from the resouces below something like twice]



### Many great resources for other approaches to learning R, and learning R for epidemiology

[Michael check]


Hadley Wickham R for Data Science book

* https://r4ds.had.co.nz/


Applied Epi:

* Home: https://appliedepi.org/
* The Epidemiologist R Handbook:  https://www.epirhandbook.com/en/
* Free self-paced tutorial series: https://appliedepi.org/tutorial/
  * need free account 
  * basics:  https://tutorials.appliedepi.org/app/basics
  * data preparation
  * data cleaning
  * basics of data visualization in R
  * summary tables


CCLHDME 2019 workshop

R Studio cheat sheets:  https://www.rstudio.com/resources/cheatsheets/

Google

YouTube

Stackoverflow   -- https://stackoverflow.com/



https://community.rstudio.com/









