---
title: "CCLHDME R Introduction"
output: html_document
date: "2022-07-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


[clean/short/simple by Tuesday mid-day]
[Michael add info about SECURITY...]
[clean up our own stuff for easy installation -- email simple guide before meeting]


mcs - purpose of this markdown is to describe the plan for the workshop, share training material and links, and ???



mcs - get started actually using R in a clean and **simple** way for real-world epidemiology and data management tasks



(except, a quick detor about objects)

learn by seeing and doing approach
Many great resources for other approaches to learning R, and learning R for epidemiology






THEN, go back and learn some fundamentals, principle and options
many ways to get data in
multiple approach to "paths" and locations of data and otehr files
many ways to load packages
base R versus "tidyverse" (versus other things like data.table)
many ways to make charts (base, ggplot2, plotly)
infinite ways to analyze data


(Future session) AND, may build sequentially on this staring point:
functions
markdown
tables
advanced data manipulation
epidemiologic statistics and modeling
shiny
shiny with GIS
etc...


[Jaspo]
what is R
installing R
Visit this website https://www.r-project.org/ and download the latest version of R suitable for your computer.


what is R Studio
installing R Studio
Visit this website https://rstudio.com/products/rstudio/download/ and download the latest free Desktop version of RStudio suitable for your computer.

[rstudio screen shot here]
walk through R studio (and return to this ...  Environment; Help, Chart, View, later Git)



[end Jaspo]



simplest uses of R

@ do some stuff in the console


introduce, define, and show: Script files (IDE)




@ script 1

simplest objects R

* atomic vectors (vector of length 1)
  * data types
    * numeric
    * character
    * logical (Boolean)

@ script 2


what are R packages
installing R packages

a few packages...

dplyr
ggplot2
readr
tidyr
summarytools
(outbreaks)
("epikit") or epitools or something else
lubridate


for later:
... for reading SAS files
readr
tmap
tables...


@ script 3














* more R object types
  *  data.frames (tibbles)
  *  lists (lot covered today)

* more about R objects
  * classes
  
* operators
  [link to operator page]



```{r}
plot(1:20)
```












Training materials and resources
https://appliedepi.org/
https://appliedepi.org/tutorial/

The Epidemiologist R Handbook
https://www.epirhandbook.com/en/


https://appliedepi.org/epirhandbook/

https://tutorials.appliedepi.org/app/basics

hadly data science book
cheat sheets
ytube
tidyverse documentation





Introduce markdown 
define etc
show markdown version of "script 3"
[markdown file]





example of rmarkdown report
example of shiny app


issues of R security


Future topics
more data manipulation
R markdown
R shiny
GIS 





