---
title: "CCLHDME R Introduction"
output: html_document
date: "2022-07-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


[clean/short/simple by Tuesday mid-day]

[Michael add info about SECURITY...]

[clean up our own stuff for easy installation -- email simple guide before meeting]

[We should revert our RStudios back to the default settings]


###  This R markdown document is a guide and tool for the workshop, and a way to share learning materials and other resources

###  The learning approach is watching leader live demonstrations, then participants doing the same things live.  Staring with immediately using R and R Studio as basically as possible for practical things, then cycle "backwards" for some concepts and options, then applying/using those, and repeat. And, pointing out and linking to additional learning resources and tools at each stage. Sound good? 



[Jaspo]

* what is R
* installing R
* Visit this website https://www.r-project.org/ and download the latest version of R suitable for your computer.
* install R


* what is R Studio
* installing R Studio
* Visit this website https://rstudio.com/products/rstudio/download/ and download the latest free Desktop version of RStudio suitable for your computer.
* install R Studio

* rstudio screen shot here

* walk through R studio (and repeatedly return to this ...  Environment; Help, Chart, View, later Git)
* overview of the R studio workspace

[end Jaspo]






@ script 1a (console)

* using R as a calculator
* script files

@ script 1b (console, then in script file)

* simplest **objects** in R
  * vectors of length 1 (atomic vectors) 
  * vectors
  * data types
    * numeric
    * character
    * logical (Boolean)
* math with simple objects    


@ script 2 (console, then in script file)

* reading some data
* using a script file to save code and start a "program"
* looking at that data
  * names()
  * table()
  * plot with base R
  * etc.
  
what are R packages

installing R packages

@ script 3 (script file)

* install a couple packages
* read some data
* wrangle some data
* ggplot some data



Some important concepts and options - Session 1

* loading packages
* reading data
* R (mathmetical) operators
* working with paths
* base R
  * indexing
  * charts
* tidyverse
  * dplyr()
    * verbs
    * pipes
  * ggplot2()
  * readr()
  * lubridate()
* using an R Project
* thousands of other packages
  * summarytools()  - easy frequencies, "cross-tables", and summary of a data.frame
  * outbreaks() - line-listed data from a bunch of real outbreaks
  * epikit() or epitools() or something else 


Some important concepts and options - Future sessions

* more R object types
  *  data.frames and tibbles
  *  lists 
  *  class()
* functions
* markdown
  * https://skylab.cdph.ca.gov/communityBurden/_w_21945fed/xMDA/2020_Excess_Mortality.html
* tables
* advanced data manipulation
* epidemiologic statistics and modeling
* shiny
* shiny with GIS
* thousands of other packages
  * reading data
    * EXCELL
     * SAS  
  * Shiny - interactive visualizations and web sites
  * visual display
  * epidemiology
  * disease surveillance
  * outbreaks
  * GIS
    * tmap()
  * data analysis and modeling
  * tables
* GIT


```{r}
height <- rnorm(n=100, mean= 60, sd = 6)
weight <- height*1.2 +  rnorm(n=100, mean= 10, sd = 6)
hist(height, breaks=50)
plot(height, weight)
```


Many great resources for other approaches to learning R, and learning R for epidemiology


Training materials and resources
https://appliedepi.org/
https://appliedepi.org/tutorial/

The Epidemiologist R Handbook
https://www.epirhandbook.com/en/


https://appliedepi.org/epirhandbook/

https://tutorials.appliedepi.org/app/basics

hadly data science book

cheat sheets

ytube

tidyverse documentation









